Modeling the Circumnuclear Mediums and Accetion Rates of Quiescent Supermassive Black Holes

### 2015-01-27 ms.tex
smbh heating

{\bf AG: Note that I obtain Compton heating which is a factor of 2
  higher--although power law dependences look right to me. Note that
  the Compton heating will depend on radius--I use rs in my own
  estimate.  BDM: if so, the please update the numbers!  }
The importance of Compton heating can be estimated by assuming the SMBH
radiates with a luminosity $L = \epsilon_{\rm rad} \dot{M}c^{2}$, where
$\epsilon_{\rm rad}$ is the radiative efficiency and where $\dot{M}$ is
estimated from equation (\ref{eq:mdot_analytic}).  Then using
equations (\ref{eq:rs_simple}), (\ref{eq:mdot_analytic}),
(\ref{eq:rhors}), (\ref{eq:rhostarrs}) we calculate that from equation
(\ref{eq:vw_eff}) that the effective Compton heating rate at the stagnation radius is given by
\begin{align} v_{w}^{\rm C} \simeq
  \begin{cases} 13 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.38}v_{500}^{-1.4}\,{\rm
km\,s^{-1}} &, \text{core}\\ 20 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.24}v_{500}^{-0.7}\,{\rm
km\,s^{-1}} &, \text{cusp},
  \end{cases}
  \label{eq:vC}
\end{align} where $T_{C,9} = T_{C}/10^{9}$ K and $\epsilon_{-2} =
\epsilon_{\rm rad}/0.01 \sim 1$ is the maximum radiative efficiency for a thick
accretion flow (e.g.~\citealt{Narayan&Yi95}).  If heating
were dominated by Compton heating, i.e. $\tilde{v}_{w} = v_{w}^{C}$, then equation (\ref{eq:vC}) can be solved for $\tilde{v}_{\rm w}$ to find
\begin{align} v_{w}^{\rm C} \simeq
  \begin{cases} 106 \eta_{0.02}^{0.21}T_{\rm
C,9}^{0.21}\epsilon_{-2}^{0.21} M_{\bullet,8}^{0.16}\,{\rm km\,s^{-1}}
&, \text{core}\\ 76 \eta_{0.02}^{0.29}T_{\rm
C,9}^{0.29}\epsilon_{-2}^{0.29} M_{\bullet,8}^{0.14}\,{\rm km\,s^{-1}}
&, \text{cusp},
  \end{cases}
  \label{eq:vC2}
\end{align} 
The fact that $v_{w}^{\rm C} \lesssim 200$ km s$^{-1}$ for $\eta \lesssim 1$ shows that Compton heating alone is insufficient to stabilize the flow against cooling
instability, i.e. $v_{w}^{C} < v_{\rm CI}$ (eq.~[\ref{eq:cooling2}]).

In contrast to radiative feedback, kinetic feedback occurs in the form
of an mass outflow or jet from the viscinity of the black hole, which
deposits its energy at much larger radii.  We assume that the total
heating rate from the SMBH is $\epsilon_{\rm j} \dot{M}c^{2}$, where
$\epsilon_{\rm j} < 1$ is an efficiency factor.  If this energy is
deposited uniformly out to a radius $r_{\rm heat}$ and volume $V_{\rm
heat}$, then the local heating rate $e = \epsilon_{\rm
j}\dot{M}c^{2}/V_{\rm heat}$ results in an effective wind heating rate
of
\begin{eqnarray} v_{w}^{\bullet} &\approx& \left(\frac{2t_{\rm
h}\epsilon_{\rm j}\dot{M}c^{2}}{V_{\rm heat}\eta
\rho_{\star}}\right)^{1/2} \sim \left(\frac{\epsilon_{\rm
j}\dot{M}c^{2}t_{\rm h}}{\eta M_{\rm enc}}\right)^{1/2} \nonumber \\
&\approx& c\epsilon_{\rm j}^{1/2}\left(\frac{r_{\rm s}}{r_{\rm
heat}}\right)^{1-\Gamma/2},
\end{eqnarray} where we have used the fact that $\dot{M} \approx \eta
M_{\rm enc}(r_{\rm s})/t_{\rm h}$ and that $M_{\rm enc}(r) \propto
M_{\bullet}(r/r_{\rm soi})^{2-\Gamma}$ for radii interior to the outer
break radius (eq.~[\ref{eq:rhostar}]).  Thus for $r_{\rm heat} \sim
r_{\rm s}$, even a small heating efficiency $\epsilon_{\rm j} \sim
10^{-6}$ is sufficient for $v_{w}^{\bullet}$ to exceed other
sources of non-accretion powered heating.  However, if the jet
distributes its energy on much larger scales comparable to the galaxy
$r_{\rm heat} \sim $ kpc $\sim 10^{3}r_{\rm s}$ then the heating rate
would be much smaller.

For a fixed heating radius $r_{\rm heat}$ we just see that $v_{\rm
  w,0}^{\bullet} \propto r_{\rm s}^{\beta/2} \propto
M_{\bullet}^{1-\Gamma/2}v_{w}^{\Gamma-2}$ is proportional to
$v_{w}$ to a negative power, where the last proportionality makes
use of equation (\ref{eq:stag_simple}).  As in the case of Compton
heating, SMBH kinetic heating, when dominant, may thus
``self-regulate".  However, unlike in the case of local stellar wind
heating, SMBH heating may not be thermally stable, since the central
accretion rate does not respond instantaneously to a change in the
thermodynamic properties of the gas on larger scales.  Also note that
$v_{w}^{\bullet}$ is proportional to a positive power of SMBH
mass $\propto M_{\bullet}^{1(0.5)}$ for core(cusp) galaxies,
respectively.  This is dependence may have important implications for
the predicted dependence of X-ray luminosity on SMBH mass
($\S\ref{sec:mdot}$).

>>>>

{\bf AG: Note that I obtain Compton heating which is a factor of 2
  higher--although power law dependences look right to me. Note that
  the Compton heating will depend on radius--I use rs in my own
  estimate.  BDM: if so, the please update the numbers!  }
The importance of Compton heating can be estimated by assuming the SMBH
radiates with a luminosity $L = \epsilon_{\rm rad} \dot{M}c^{2}$, where
$\epsilon_{\rm rad}$ is the radiative efficiency and where $\dot{M}$ is
estimated from equation (\ref{eq:mdot_analytic}).  Then using
equations (\ref{eq:rs_simple}), (\ref{eq:mdot_analytic}),
(\ref{eq:rhors}), (\ref{eq:rhostarrs}) we calculate that from equation
(\ref{eq:vw_eff}) that the effective Compton heating rate at the stagnation radius is given by
\begin{align} v_{w}^{\rm C} \simeq
  \begin{cases} 13 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.38}v_{500}^{-1.4}\,{\rm
km\,s^{-1}} &, \text{core}\\ 20 \eta_{0.02}^{0.5}T_{\rm
C,9}^{0.5}\epsilon_{-2}^{0.5} M_{\bullet,8}^{0.24}v_{500}^{-0.7}\,{\rm
km\,s^{-1}} &, \text{cusp},
  \end{cases}
  \label{eq:vC}
\end{align} where $T_{C,9} = T_{C}/10^{9}$ K and $\epsilon_{-2} =
\epsilon_{\rm rad}/0.01 \sim 1$ is the maximum radiative efficiency for an advective
accretion flow expected to produce high $T_{\rm C}$ emission (e.g.~\citealt{Narayan&Yi95}).  If heating is provided by Compton heating alone, i.e. $\tilde{v}_{w} = v_{w}^{C}$, then equation (\ref{eq:vC}) can be solved for $\tilde{v}_{\rm w}$ to find
\begin{align} v_{w}^{\rm C} \simeq
  \begin{cases} 106 \eta_{0.02}^{0.21}T_{\rm
C,9}^{0.21}\epsilon_{-2}^{0.21} M_{\bullet,8}^{0.16}\,{\rm km\,s^{-1}}
&, \text{core}\\ 76 \eta_{0.02}^{0.29}T_{\rm
C,9}^{0.29}\epsilon_{-2}^{0.29} M_{\bullet,8}^{0.14}\,{\rm km\,s^{-1}}
&, \text{cusp},
  \end{cases}
  \label{eq:vC2}
\end{align} 
The fact that $v_{w}^{\rm C} \lesssim 200$ km s$^{-1}$ for physically realized values of $\eta \lesssim 1$ shows that Compton heating alone is insufficient to stabilize the flow against cooling
instability, i.e. $v_{w}^{C} < v_{\rm CI}$ (eq.~[\ref{eq:cooling2}]).

Kinetic feedback occurs from an outflow of mass or energy  from the viscinity of the black hole in the form of a disk wind or jet, which deposits its energy across a much wider range of radii.  If the total outflow power is proportional to the SMBH accretion rate, $L_{\rm j} = \epsilon_{\rm j} \dot{M}c^{2}$, where $\epsilon_{\rm j} < 1$ is an efficiency factor.  Assuming that this energy is deposited uniformly out to a radius $r_{\rm heat}$ and volume $V_{\rm heat} \propto r_{\rm heat}^{3}$, then the volumetric heating rate $e = \epsilon_{\rm j}\dot{M}c^{2}/V_{\rm heat}$ results in an effective wind velocity at the stagnation radius given by
\begin{eqnarray} v_{w}^{\bullet} &\approx& \left(\frac{2t_{\rm
h}\epsilon_{\rm j}\dot{M}c^{2}}{V_{\rm heat}\eta
\rho_{\star}}\right)^{1/2} \sim \left(\frac{\epsilon_{\rm
j}\dot{M}c^{2}t_{\rm h}}{\eta M_{\rm enc}}\right)^{1/2} \nonumber \\
&\approx& 300\,{\rm km\,s^{-1}}\,\left(\frac{\epsilon_{\rm j}}{10^{-6}}\right)^{1/2}\left(\frac{r_{\rm s}}{r_{\rm
heat}}\right)^{1-\Gamma/2},
\end{eqnarray} where we have used the fact that $\dot{M} \approx \eta
M_{\rm enc}(r_{\rm s})/t_{\rm h}$ and that $M_{\rm enc}(r) \propto
M_{\bullet}(r/r_{\rm soi})^{2-\Gamma}$ for radii interior to the outer
break radius (eq.~[\ref{eq:rhostar}]).  Thus for $r_{\rm heat} \sim
r_{\rm s}$, even a modest heating efficiency $\epsilon_{\rm j} \gtrsim
10^{-5}$ is sufficient for $v_{w}^{\bullet}$ to exceed other
sources of non-accretion powered heating.  However, if the jet distributes its energy on much larger scales comparable to the size of the galaxy, i.e. $r_{\rm heat} \gtrsim $ kpc $\gtrsim 10^{3}r_{\rm s}$ then the heating rate would be much less significant, even for an efficient jet ($\epsilon_{\rm j} \sim 1$).

Also note that for a fixed $r_{\rm heat}$, the effective heating rate $v_{\rm  w}^{\bullet} \propto r_{\rm s}^{1-\Gamma/2} \propto M_{\bullet}^{1-\Gamma/2}v_{w}^{\Gamma-2}$ is proportional to $v_{w}$ to a negative power (where we have used of eq.~[\ref{eq:stag_simple}]).  This shows that SMBH kinetic heating, if dominant, could thus `self-regulate' insofar as a lower heating rate results in a higher accretion rate, which in turn results in stronger jet feedback.  Importantly, however, unlike in the case of local heating sources such as stellar winds, even a dominant SMBH heating may not allow for a thermally-stable steady flow because the central accretion rate does not respond instantaneously to changes in the
thermodynamic properties of the gas on larger scale.  We also note here that the kinetic heating rate is proportional to the SMBH mass to a positive power, $v_{w}^{\bullet} \propto M_{\bullet}^{1(0.5)}$ for core(cusp) galaxies, respectively. This dependence could have important implications for the dependence of the average nuclear X-ray luminosity of quiescent galaxies on the SMBH mass ($\S\ref{sec:mdot}$).
